---
title: "Adjusting_populations_pt4"
output: html_document
date: "2024-07-28"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



Create a table for HSC population pre-enrichment (taken from supplementary table S2B) for pt4 BL
```{r}
library(Seurat)
library(tidyverse)
```
```{r}
x<-c("HSC", "MPP", "LMPP", "CMP", "GMP", "MEP", "MLP", "B-NK", "Lin-mature","Precursors")
y<-c(4.0, 0.1, 0.04, 84.8, 2.0,9.1, 0.01, 0)
y<- y*0.305
y<-c(y, 69, 1)
y
x_name<-"population"
y_name<-"proportion"
df<-data.frame(x,y)
names(df)<-c(x_name, y_name)
head(df, 5)


```
Calclulate the proportions after enrichment from the transcriptome metadata.
```{r}
obj<-readRDS("data/4_Seurat_obj_CNA_referenceannotations.rds")
pt4<-subset(x=obj, subset=Patient=='pt4')
```
```{r}
pt4_BL<-subset(x=pt4, subset=pt_status =='pt4 BL')
pt4_BL<-pt4_BL@meta.data
```

```{r}
x_df<-pt4$Population
y_df<-pt4$pt_status
df_enrich<-data.frame(x_df, y_df)
head(df_enrich, 5)
df_BL<-subset(df_enrich, y_df=="pt4 BL")
```
```{r}
library(tidyverse)
table <-table(df_BL$y_df, df_BL$x_df) %>% prop.table
table
```
```{r}
z<-c(18.1, 0.2, 0, 79.0, 0.2, 0.7, 0, 0, 1.6, 0.1) #Take proprotions from table
a<-c("HSC", "MPP", "LMPP", "CMP", "GMP", "MEP", "MLP", "B-NK", "Lin-mature","Precursors")
df_RNA<-data.frame(a,z)
a_name<-"population"
z_name<-"proportion_RNA"
names(df_RNA)<-c(a_name, z_name)
df_RNA$proportion_RNA<-df_RNA$proportion_RNA/100
head(df_RNA, 5)
head(df, 5)
```
```{r}
library(tidyverse)
df_com<-inner_join(df, df_RNA, by = "population")
```
```{r}
head(df_com, 5)
```
```{r}
df_com$difference<-df_com$proportion-df_com$proportion_RNA
df_com$difference
```

```{r}
 #Check how the populations are written to match
df_com # take proportions preenrichment, but only from populations represented in the RNA df
```
```{r}
proportions<-c(0.2586, 0.006, 0.905, 0.003, 0.001, 0.03,0.0122)
Population<-c("CMP", "GMP", "Lin-Mature", "MPP", "Precursor", "MEP","HSC")
fraction_sample<-data.frame(Population, proportions)
fraction_sample
joined<-pt4_BL%>%left_join(fraction_sample, by = 'Population')%>% group_by(Population) %>% sample_n(n()*first(proportions))%>%ungroup ()


  #select(-proportions)
joined

  
```
```{r}
head(joined)
```

```{r}
table <-table(joined$Population) %>% prop.table
table
```
The following steps filters the cells from patient 4 BL based on the cells selected by the sampling step. The column with the unique identifier was lost. To circumvent that I'm selecting for a value which is unique to the cells NCount_gene. It seems to work because the number of rows is the same in the sampled dataset as in the filtered seurat object.
```{r}
selected_ids<-joined$Sample_well
unique_identifier<-joined$nCount_gene
unique(selected_ids)
duplicated(selected_ids)
duplicated(unique_identifier)
pt4_BL_seurat<-subset( x = pt4, subset= pt_status == "pt4 BL")
nrow(pt4_BL_seurat@meta.data)
pt4_BL_seurat_filter<-subset(x= pt4_BL_seurat, subset = nCount_gene %in% unique_identifier )
nrow(pt4_BL_seurat_filter@meta.data)
```
This filtered Seurat object should now be ready to be projected on the reference.
```{r}
library(Seurat)
library(tidyverse)
library(symphony)
library(ggpubr)
library(patchwork)
library(AUCell)
library(jcolors)
library(BoneMarrowMap)
library(RColorBrewer)
```
Set up the reference map.
```{r}
projection_path = 'results/'
batchvar<-'pt_status'
# Load Symphony reference
ref <- readRDS(paste0(projection_path, 'BoneMarrow_RefMap_SymphonyRef.rds'))
# Set uwot path for UMAP projection
ref$save_uwot_path <- paste0(projection_path, 'BoneMarrow_RefMap_uwot_model.uwot')
#Visualize the Reference BM
ReferenceSeuratObj <- create_ReferenceObject(ref)
DimPlot(ReferenceSeuratObj, reduction = 'umap', group.by = 'CellType_Annotation_formatted', 
        raster=FALSE, label=TRUE, label.size = 4)
#Visualize the Reference BM
ReferenceSeuratObj <- create_ReferenceObject(ref)
DimPlot(ReferenceSeuratObj, reduction = 'umap', group.by = 'CellType_Annotation_formatted', 
        raster=FALSE, label=TRUE, label.size = 4)
#Map the data, mapping done using Symphony (Kang et al, 2021)
mapped_BL <- map_Query(
  exp_query = pt4_BL_seurat_filter@assays$RNA$counts, 
  metadata_query = pt4_BL_seurat_filter@meta.data,
  ref_obj = ref,
  vars = batchvar
)
#Exclude cells with a high mapping error
mapped_BL<-mapped_BL%>% calculate_MappingError(., reference=ref, MAD_threshold = 2.5)
#Remove cells with a high mapping error (optional)
mappedQC_BL <- subset( mapped_BL, mapping_error_QC == 'Pass')

#Cell Type assignment using 30 K-Nearest Neighbours
# Predict Hematopoietic Cell Types by KNN classification
mappedQC_BL <- predict_CellTypes(
  query_obj = mappedQC_BL, 
  ref_obj = ref, 
  initial_label = 'initial_CellType', # celltype assignments before filtering on mapping QC
  final_label = 'predicted_CellType'  # celltype assignments with map QC failing cells assigned as NA
) 

#Predict Pseudotime values by k-nearest neighbours
mappedQC_BL <- predict_Pseudotime(
  query_obj = mappedQC_BL,
  ref_obj = ref,
  initial_label = 'initial Pseudotime',
  final_label = 'predicted_Pseudotime')
FeaturePlot(mappedQC_BL, features = c ('predicted_Pseudotime'))
```

```{r}
FeaturePlot(mappedQC_BL, features = c ('predicted_Pseudotime'))
```
```{r}
DimPlot(mappedQC_BL, reduction = 'umap', group.by = c('predicted_CellType'), raster=FALSE, label=TRUE, label.size = 4)
```


Repeat the same for Pt4 at relapse.
```{r}
x<-c("HSC", "MPP", "LMPP", "CMP", "GMP", "MEP", "MLP", "B-NK", "Lin-mature","Precursors")
y<-c(7.2,0.1, 0.25, 73.2, 6.2, 12.7, 0.3, 0.1)
y<- y*0.74
y<-c(y, 23.5, 2.5)
y<-y/100
x_name<-"population"
y_name<-"proportion"
df<-data.frame(x,y)
names(df)<-c(x_name, y_name)
head(df, 5)
```

```{r}
x_df<-pt4$Population
y_df<-pt4$pt_status
df_enrich<-data.frame(x_df, y_df)
head(df_enrich, 5)
df_REL<-subset(df_enrich, y_df=="pt4 REL")
```
```{r}
table <-table(df_REL$y_df, df_REL$x_df) %>% prop.table
table
```
```{r}
z<-c(0.16,0.01,0.64,0.03, 0.014, 0, 0, 0.033, 0.102, 0.004)
a<-c("HSC", "MPP", "CMP", "GMP", "MEP", "MLP", "B-NK", "Lin-mature","Precursors", "LMPP")
df_RNA<-data.frame(a,z)
a_name<-"population"
z_name<-"proportion_RNA"
names(df_RNA)<-c(a_name, z_name)
head(df_RNA, 5)
df_RNA
```
```{r}
df_com<-inner_join(df, df_RNA, by = "population")
```
```{r}
head(df_com,5)
```
```{r}
df_com$difference<-df_com$proportion-df_com$proportion_RNA
df_com$difference
df_com
```
```{r}
#proportions<-df_com$proportion
#Population<-df_com$population
pt4_REL<-subset(x=pt4, subset=pt_status =='pt4 REL')
pt4_REL<-pt4_REL@meta.data
table(pt4_REL$Population) #adjust the names of the populations accordingly, otherwise the joining doesn't work
```
```{r}
proportions<-c(0.54, 0.05, 0.053, 0.000, 0.23, 0.025,0, 0.09,0, 0.0002)
Population<-c("CMP", "GMP", "HSC","MPP","Lin-Mature", "Precursor", "B-NK", "MEP", "MLP", "LMPP") #only populations represented in the RNA data
fraction_sample<-data.frame(Population, proportions)
fraction_sample
joined<-pt4_REL%>%left_join(fraction_sample, by = 'Population')%>% 
  group_by(Population)%>% sample_n(n()*first(proportions))%>%ungroup%>%
  select(-proportions)
head(joined,5)
```
```{r}
table <-table(joined$Population) %>% prop.table
table
```
The proportions do not match exactly. I guess that's because Lin-Mature cells for examples are underrepresented in the dataset so you cannot match exactly!?
I'll add the metadata in the same very unelegant way I did for the BL dataset.

```{r}
selected_ids<-joined$Sample_well
unique_identifier<-joined$nCount_gene
unique(selected_ids)
duplicated(selected_ids)
duplicated(unique_identifier)
pt4_REL_seurat<-subset( x = pt4, subset= pt_status == "pt4 REL")
nrow(pt4_REL_seurat@meta.data)
pt4_REL_seurat_filter<-subset(x= pt4_REL_seurat, subset = nCount_gene %in% unique_identifier )
nrow(pt4_REL_seurat_filter@meta.data)
```
Project the reduced cell number in the reference.
```{r}
#Map the data, mapping done using Symphony (Kang et al, 2021)
mapped_REL <- map_Query(
  exp_query = pt4_REL_seurat_filter@assays$RNA$counts, 
  metadata_query = pt4_REL_seurat_filter@meta.data,
  ref_obj = ref,
  vars = batchvar
)
#Exclude cells with a high mapping error
mapped_REL<-mapped_REL%>% calculate_MappingError(., reference=ref, MAD_threshold = 2.5)
#Remove cells with a high mapping error (optional)
mappedQC_REL <- subset( mapped_REL, mapping_error_QC == 'Pass')

#Cell Type assignment using 30 K-Nearest Neighbours
# Predict Hematopoietic Cell Types by KNN classification
mappedQC_REL <- predict_CellTypes(
  query_obj = mappedQC_REL, 
  ref_obj = ref, 
  initial_label = 'initial_CellType', # celltype assignments before filtering on mapping QC
  final_label = 'predicted_CellType'  # celltype assignments with map QC failing cells assigned as NA
) 

#Predict Pseudotime values by k-nearest neighbours
mappedQC_REL <- predict_Pseudotime(
  query_obj = mappedQC_REL,
  ref_obj = ref,
  initial_label = 'initial Pseudotime',
  final_label = 'predicted_Pseudotime')
FeaturePlot(mappedQC_REL, features = c ('predicted_Pseudotime'))
```


```{r}
DimPlot(mappedQC_REL, reduction = 'umap', group.by = c('predicted_CellType'), raster=FALSE, label=TRUE, label.size = 4)
```
```{r}
pt4_BL_REL<-merge(mappedQC_BL, mappedQC_REL)
pt4_celltype<-get_Composition(query_obj = pt4_BL_REL, donor_key = "Sample", celltype_label="predicted_CellType", return_type = 'proportion')
heatmap4<-pt4_celltype%>%column_to_rownames('Sample')%>%data.matrix()%>%ComplexHeatmap::Heatmap()
heatmap4

pt4_celltype<-get_Composition(query_obj = pt4_BL_REL, donor_key = "Sample", celltype_label="predicted_CellType", return_type = 'proportion')
heatmap4<-pt4_celltype%>%column_to_rownames('Sample')%>%data.matrix()%>%ComplexHeatmap::Heatmap(row_order = c("REL", "BL"))
heatmap4
```
Filter for the dominant clone only
```{r}
data_pt4_BL_REL<-pt4_BL_REL@meta.data
data4<-mutate(data_pt4_BL_REL, dominant = if_else(
  pt_status == 'pt4 BL'&clone.y %in% c("SIA", "SIAC", "SIAJ")|
pt_status == 'pt4 REL' & clone.y %in% c("SIA_20_1", "SIA_20_7"), "1","0"))
#Check whether it worked
table(data4$dominant, data4$clone.y, data4$Sample)
```
```{r}
#Add metadata back to Seurat object
pt4_BL_REL<-AddMetaData(object = pt4_BL_REL, metadata = data4, col.name = 'dominant')
pt4_BL_REL <- subset(x=pt4_BL_REL, dominant == "1")
pt4_celltype<-get_Composition(query_obj = pt4_BL_REL, donor_key = "Sample", celltype_label="predicted_CellType", return_type = 'proportion')
heatmap4<-pt4_celltype%>%column_to_rownames('Sample')%>%data.matrix()%>%ComplexHeatmap::Heatmap(row_order= c("REL", "BL"))
heatmap4
```
```{r}
pt4_BL_REL[["RNA"]]<-JoinLayers(pt4_BL_REL[["RNA"]])
mapped_QC <- map_Query(
  exp_query = pt4_BL_REL@assays$RNA$counts, 
  metadata_query = pt4_BL_REL@meta.data,
  ref_obj = ref,
  vars = batchvar
)
#Exclude cells with a high mapping error
mapped_REL<-mapped_QC%>% calculate_MappingError(., reference=ref, MAD_threshold = 2.5)
#Remove cells with a high mapping error (optional)
mappedQC <- subset( mapped_QC, mapping_error_QC == 'Pass')

#Cell Type assignment using 30 K-Nearest Neighbours
# Predict Hematopoietic Cell Types by KNN classification
mappedQC <- predict_CellTypes(
  query_obj = mappedQC, 
  ref_obj = ref, 
  initial_label = 'initial_CellType', # celltype assignments before filtering on mapping QC
  final_label = 'predicted_CellType'  # celltype assignments with map QC failing cells assigned as NA
) 
umap_plot<-DimPlot(mapped_QC, reduction = 'umap', group.by = c('predicted_CellType'), raster=FALSE, label=TRUE, label.size = 4)
print(umap_plot)
ggsave("UMAP Pt4 on reference.pdf", plot = umap_plot, path = "results/Pt4/results/Figures", width =20, height = 30, unit = "cm")
```
```{r}
umap_data <-as.data.frame(umap_plot$data)
metadata<-pt4_BL_REL@meta.data
umap_data<-data.frame(names = row.names(umap_data), umap_data)
metadata<-data.frame(names = row.names(metadata), metadata)
merged_umap <- full_join(umap_data, metadata, by = "names")
pt_status == 'pt4 BL'&clone.y %in% c("SIA", "SIAC", "SIAJ")|
    pt_status == 'pt14 REL' & clone.y %in% c("SIA_20_1", "SIA_20_7")~ "1"
merged_umap<-merged_umap%>%mutate(highlight = case_when(
 clone.y == c("SIA", "SIAC", "SIAJ") & Sample == "BL"~ "1",
  clone.y ==  c("SIA_20_1", "SIA_20_7") & Sample == "REL" ~ "2", 
  clone.y!="SIA"|clone.y !="SIAJ"|clone.y !="SIAC"|clone.y !="SIA_20_1"|clone.y !="SIA_20_7" ~ "3"))
sum(is.na(merged_umap$highlight))
merged_umap<-merged_umap%>%arrange(desc(highlight)) #reorders the df so that colored points are on top of the grey ones

cols <- c("1" = "darkorange3", "2"="darkolivegreen4", "3"="darkgrey", "NA" = "darkgrey")
umap_highlight<-ggplot(merged_umap, aes(x = umap_1, y = umap_2))+
  geom_point(aes(color = highlight), size = 0.4)+
  scale_color_manual(values = cols)+
  labs(color = "Cells")+
  theme_minimal()+
  theme(legend.position = "right")
ggsave("UMAP Pt4 on reference_highlight.pdf", plot = umap_highlight, path = "results/Pt4/results/Figures", width =20, height = 30, unit = "cm")
print(umap_highlight)
```

