---
title: "Adjusting_populations_pt14"
output: html_document
date: "20214-07-28"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



Create a table for HSC population pre-enrichment (taken from supplementary table S2B) for pt14 BL
```{r}
library(Seurat)
library(tidyverse)
```
```{r}
x<-c("HSC", "MPP", "LMPP", "CMP", "GMP", "MEP", "MLP", "B-NK", "Lin-mature","Precursors")
y<-c(19.3, 36.6, 7.4, 12.5, 10.4,13.53, 0, 0.3)
y<- y*0.02
y<-c(y, 95, 3)
y
x_name<-"population"
y_name<-"proportion"
df<-data.frame(x,y)
names(df)<-c(x_name, y_name)
head(df, 5)


```
Calclulate the proportions after enrichment from the transcriptome metadata.
```{r}
obj<-readRDS("data/4_Seurat_obj_CNA_referenceannotations.rds")
pt14<-subset(x=obj, subset=Patient=='pt14')
```
```{r}
pt14_BL<-subset(x=pt14, subset=pt_status =='pt14 BL')
pt14_BL<-pt14_BL@meta.data
```

```{r}
x_df<-pt14$Population
y_df<-pt14$pt_status
df_enrich<-data.frame(x_df, y_df)
head(df_enrich, 5)
df_BL<-subset(df_enrich, y_df=="pt14 BL")
```
```{r}
library(tidyverse)
table <-table(df_BL$y_df, df_BL$x_df) %>% prop.table
table
```
```{r}
z<-c(2.4, 14.8, 4, 8, 6, 0.8, 0, 0.4, 8.4, 45.2) #Take proprotions from table
a<-c("HSC", "MPP", "LMPP", "CMP", "GMP", "MEP", "MLP", "B-NK", "Lin-mature","Precursors")
df_RNA<-data.frame(a,z)
a_name<-"population"
z_name<-"proportion_RNA"
names(df_RNA)<-c(a_name, z_name)
df_RNA$proportion_RNA<-df_RNA$proportion_RNA/100
head(df_RNA, 5)
head(df, 5)
```
```{r}
library(tidyverse)
df_com<-inner_join(df, df_RNA, by = "population")
```
```{r}
head(df_com, 5)
```
```{r}
df_com$difference<-df_com$proportion-df_com$proportion_RNA
df_com$difference
```

```{r}
 #Check how the populations are written to match
df_com # take proportions preenrichment, but only from populations represented in the RNA df
```
```{r}
proportions<-c(0.003, 0.002, 0.95, 0.001, 0.007, 0.03, 0.003,0.004,0, 0)
Population<-c("CMP", "GMP", "Lin-Mature", "LMPP", "MPP", "Precursor", "MEP","HSC", "MLP", "B-NK")
fraction_sample<-data.frame(Population, proportions)
fraction_sample
joined<-pt14_BL%>%left_join(fraction_sample, by = 'Population')%>% 
  group_by(Population) %>% sample_n(n()*first(proportions))%>%ungroup%>%
  select(-proportions)
joined

  
```
```{r}
head(joined)
```

```{r}
table <-table(joined$Population) %>% prop.table
table
```
The following steps filters the cells from patient 14 BL based on the cells selected by the sampling step. The column with the unique identifier was lost. To circumvent that I'm selecting for a value which is unique to the cells NCount_gene. It seems to work because the number of rows is the same in the sampled dataset as in the filtered seurat object.
```{r}
selected_ids<-joined$Sample_well
unique_identifier<-joined$nCount_gene
unique(selected_ids)
duplicated(selected_ids)
duplicated(unique_identifier)
pt14_BL_seurat<-subset( x = pt14, subset= pt_status == "pt14 BL")
nrow(pt14_BL_seurat@meta.data)
pt14_BL_seurat_filter<-subset(x= pt14_BL_seurat, subset = nCount_gene %in% unique_identifier )
nrow(pt14_BL_seurat_filter@meta.data)
```
This filtered Seurat object should now be ready to be projected on the reference.
```{r}
library(Seurat)
library(tidyverse)
library(symphony)
library(ggpubr)
library(patchwork)
library(AUCell)
library(jcolors)
library(BoneMarrowMap)
library(RColorBrewer)
```
Set up the reference map.
```{r}
projection_path = 'results/'
batchvar<-'pt_status'
# Load Symphony reference
ref <- readRDS(paste0(projection_path, 'BoneMarrow_RefMap_SymphonyRef.rds'))
# Set uwot path for UMAP projection
ref$save_uwot_path <- paste0(projection_path, 'BoneMarrow_RefMap_uwot_model.uwot')
#Visualize the Reference BM
ReferenceSeuratObj <- create_ReferenceObject(ref)
DimPlot(ReferenceSeuratObj, reduction = 'umap', group.by = 'CellType_Annotation_formatted', 
        raster=FALSE, label=TRUE, label.size = 14)
#Visualize the Reference BM
ReferenceSeuratObj <- create_ReferenceObject(ref)
DimPlot(ReferenceSeuratObj, reduction = 'umap', group.by = 'CellType_Annotation_formatted', 
        raster=FALSE, label=TRUE, label.size = 14)
#Map the data, mapping done using Symphony (Kang et al, 2021)
mapped_BL <- map_Query(
  exp_query = pt14_BL_seurat_filter@assays$RNA$counts, 
  metadata_query = pt14_BL_seurat_filter@meta.data,
  ref_obj = ref,
  vars = batchvar
)
#Exclude cells with a high mapping error
mapped_BL<-mapped_BL%>% calculate_MappingError(., reference=ref, MAD_threshold = 2.5)
#Remove cells with a high mapping error (optional)
mappedQC_BL <- subset( mapped_BL, mapping_error_QC == 'Pass')

#Cell Type assignment using 30 K-Nearest Neighbours
# Predict Hematopoietic Cell Types by KNN classification
mappedQC_BL <- predict_CellTypes(
  query_obj = mappedQC_BL, 
  ref_obj = ref, 
  initial_label = 'initial_CellType', # celltype assignments before filtering on mapping QC
  final_label = 'predicted_CellType'  # celltype assignments with map QC failing cells assigned as NA
) 

#Predict Pseudotime values by k-nearest neighbours
mappedQC_BL <- predict_Pseudotime(
  query_obj = mappedQC_BL,
  ref_obj = ref,
  initial_label = 'initial Pseudotime',
  final_label = 'predicted_Pseudotime')
FeaturePlot(mappedQC_BL, features = c ('predicted_Pseudotime'))
```

```{r}
FeaturePlot(mappedQC_BL, features = c ('predicted_Pseudotime'))
```
```{r}
DimPlot(mappedQC_BL, reduction = 'umap', group.by = c('predicted_CellType'), raster=FALSE, label=TRUE, label.size = 14)
```


Repeat the same for Pt14 at relapse.
```{r}
x<-c("HSC", "MPP", "LMPP", "CMP", "GMP", "MEP", "MLP", "B-NK", "Lin-mature","Precursors")
y<-c(1.6,11.6, 15.6, 18.7, 46.8, 5.5, 0, 0.1)
y<- y*0.025
y<-c(y, 20.5, 77)
y<-y/100
x_name<-"population"
y_name<-"proportion"
df<-data.frame(x,y)
names(df)<-c(x_name, y_name)
head(df, 5)
```

```{r}
x_df<-pt14$Population
y_df<-pt14$pt_status
df_enrich<-data.frame(x_df, y_df)
head(df_enrich, 5)
df_REL<-subset(df_enrich, y_df=="pt14 REL")
```
```{r}
table <-table(df_BL$y_df, df_BL$x_df) %>% prop.table
table
```
```{r}
z<-c(0.024,0.148,0.04,0.080, 0.060, 0.008, 0,0.004, 0.184, 0.452)
a<-c("HSC", "MPP", "LMPP", "CMP", "GMP", "MEP", "MLP", "B-NK", "Lin-mature","Precursors")
df_RNA<-data.frame(a,z)
a_name<-"population"
z_name<-"proportion_RNA"
names(df_RNA)<-c(a_name, z_name)
head(df_RNA, 5)
```
```{r}
df_com<-inner_join(df, df_RNA, by = "population")
```
```{r}
head(df_com,5)
```
```{r}
df_com$difference<-df_com$proportion-df_com$proportion_RNA
df_com$difference
df_com
```

```{r}
pt14_REL<-subset(x=pt14, subset=pt_status =='pt14 REL')
```


```{r}
```


```{r}
pt14_REL<-pt14_REL@meta.data
proportions<-c(0.005, 0.004, 0.0117, 0.0004, 0.003, 0.205, 0.77, 0)
Population<-c("CMP", "LMPP", "GMP", "HSC","MPP","Lin-Mature", "Precursor", "B-NK") #only populations represented in the 
#RNA data
fraction_sample<-data.frame(Population, proportions)
fraction_sample
joined<-pt14_REL%>%left_join(fraction_sample, by = 'Population')%>% 
  group_by(Population)%>% sample_n(n()*first(proportions))%>%ungroup%>%
  select(-proportions)
head(joined,5)
```
```{r}
table <-table(joined$Population) %>% prop.table
table
```
The proportions do not match exactly. I guess that's because Lin-Mature cells for examples are underrepresented in the dataset so you cannot match exactly!?
I'll add the metadata in the same very unelegant way I did for the BL dataset.

```{r}
selected_ids<-joined$Sample_well
unique_identifier<-joined$nCount_gene
unique(selected_ids)
duplicated(selected_ids)
duplicated(unique_identifier)
pt14_REL_seurat<-subset( x = pt14, subset= pt_status == "pt14 REL")
nrow(pt14_REL_seurat@meta.data)
pt14_REL_seurat_filter<-subset(x= pt14_REL_seurat, subset = nCount_gene %in% unique_identifier )
nrow(pt14_REL_seurat_filter@meta.data)
```
Project the reduced cell number in the reference.
```{r}
#Map the data, mapping done using Symphony (Kang et al, 2021)
mapped_REL <- map_Query(
  exp_query = pt14_REL_seurat_filter@assays$RNA$counts, 
  metadata_query = pt14_REL_seurat_filter@meta.data,
  ref_obj = ref,
  vars = batchvar
)
#Exclude cells with a high mapping error
mapped_REL<-mapped_REL%>% calculate_MappingError(., reference=ref, MAD_threshold = 2.5)
#Remove cells with a high mapping error (optional)
mappedQC_REL <- subset( mapped_REL, mapping_error_QC == 'Pass')

#Cell Type assignment using 30 K-Nearest Neighbours
# Predict Hematopoietic Cell Types by KNN classification
mappedQC_REL <- predict_CellTypes(
  query_obj = mappedQC_REL, 
  ref_obj = ref, 
  initial_label = 'initial_CellType', # celltype assignments before filtering on mapping QC
  final_label = 'predicted_CellType'  # celltype assignments with map QC failing cells assigned as NA
) 

#Predict Pseudotime values by k-nearest neighbours
mappedQC_REL <- predict_Pseudotime(
  query_obj = mappedQC_REL,
  ref_obj = ref,
  initial_label = 'initial Pseudotime',
  final_label = 'predicted_Pseudotime')
FeaturePlot(mappedQC_REL, features = c ('predicted_Pseudotime'))
```


```{r}
DimPlot(mappedQC_REL, reduction = 'umap', group.by = c('predicted_CellType'), raster=FALSE, label=TRUE, label.size = 14)
```
```{r}
pt14_BL_REL<-merge(mappedQC_BL, mappedQC_REL)
pt14_celltype<-get_Composition(query_obj = pt14_BL_REL, donor_key = "Sample", celltype_label="predicted_CellType", return_type = 'proportion')
heatmap14<-pt14_celltype%>%column_to_rownames('Sample')%>%data.matrix()%>%ComplexHeatmap::Heatmap()
heatmap14

pt14_celltype<-get_Composition(query_obj = pt14_BL_REL, donor_key = "Sample", celltype_label="predicted_CellType", return_type = 'proportion')
heatmap14<-pt14_celltype%>%column_to_rownames('Sample')%>%data.matrix()%>%ComplexHeatmap::Heatmap(row_order = c("REL", "BL"))
heatmap14
```
Filter for the dominant clone only
```{r}
data_pt14_BL_REL<-pt14_BL_REL@meta.data
data14<-mutate(data_pt14_BL_REL, dominant = if_else(
  pt_status == 'pt14 BL'&clone.y %in% c("DNI")|
    pt_status == 'pt14 REL' & clone.y %in% c("DN_4", "DN_4_20", "DN_4_21"), "1","0"))
#Check whether it worked
table(data14$dominant, data14$clone.y, data14$Sample)
```
```{r}
#Add metadata back to Seurat object
pt14_BL_REL<-AddMetaData(object = pt14_BL_REL, metadata = data14, col.name = 'dominant')
pt14_BL_REL <- subset(x=pt14_BL_REL, dominant == "1")
pt14_celltype<-get_Composition(query_obj = pt14_BL_REL, donor_key = "Sample", celltype_label="predicted_CellType", return_type = 'proportion')
heatmap14<-pt14_celltype%>%column_to_rownames('Sample')%>%data.matrix()%>%ComplexHeatmap::Heatmap(row_order = c("REL", "BL"))
heatmap14
```
```{r}
pt14_BL_REL[["RNA"]]<-JoinLayers(pt14_BL_REL[["RNA"]])
mapped_QC <- map_Query(
  exp_query = pt14_BL_REL@assays$RNA$counts, 
  metadata_query = pt14_BL_REL@meta.data,
  ref_obj = ref,
  vars = batchvar
)
#Exclude cells with a high mapping error
mapped_REL<-mapped_QC%>% calculate_MappingError(., reference=ref, MAD_threshold = 2.5)
#Remove cells with a high mapping error (optional)
mappedQC <- subset( mapped_QC, mapping_error_QC == 'Pass')

#Cell Type assignment using 30 K-Nearest Neighbours
# Predict Hematopoietic Cell Types by KNN classification
mappedQC <- predict_CellTypes(
  query_obj = mappedQC, 
  ref_obj = ref, 
  initial_label = 'initial_CellType', # celltype assignments before filtering on mapping QC
  final_label = 'predicted_CellType'  # celltype assignments with map QC failing cells assigned as NA
) 
umap_plot<-DimPlot(mapped_QC, reduction = 'umap', group.by = c('predicted_CellType'), raster=FALSE, label=TRUE, label.size = 4)
print(umap_plot)
ggsave("UMAP Pt14 on reference.pdf", plot = umap_plot, path = "results/Pt14/results/Figures", width =20, height = 30, unit = "cm")
```
```{r}
umap_data <-as.data.frame(umap_plot$data)
metadata<-pt14_BL_REL@meta.data
umap_data<-data.frame(names = row.names(umap_data), umap_data)
metadata<-data.frame(names = row.names(metadata), metadata)
merged_umap <- full_join(umap_data, metadata, by = "names")
pt_status == 'pt14 BL'&clone.y %in% c("DNI")|
    pt_status == 'pt14 REL' & clone.y %in% c("DN_4", "DN_4_20", "DN_4_21")~ "1"
merged_umap<-merged_umap%>%mutate(highlight = case_when(
 clone.y == "DNI" & Sample == "BL"~ "1",
  clone.y == c("DN_4", "DN_4_20", "DN_4_21") & Sample == "REL" ~ "2", 
  clone.y!="DNI"|clone.y !="DN_4"|clone.y !="DN_4_20"|clone.y !="DN_21" ~ "3"))
sum(is.na(merged_umap$highlight))
merged_umap<-merged_umap%>%arrange(desc(highlight)) #reorders the df so that colored points are on top of the grey ones

cols <- c("1" = "darkorange2", "2"="darkolivegreen4", "3"="darkgrey", "NA" = "darkgrey")
umap_highlight<-ggplot(merged_umap, aes(x = umap_1, y = umap_2))+
  geom_point(aes(color = highlight), size = 0.4)+
  scale_color_manual(values = cols)+
  labs(color = "Cells")+
  theme_minimal()+
  theme(legend.position = "right")
ggsave("UMAP Pt14 on reference_highlight.pdf", plot = umap_highlight, path = "results/Pt14/results/Figures", width =20, height = 30, unit = "cm")
```
```{r}

```
