---
title: "Pt14_GSEA_dream"
output: html_document
date: "2024-06-16"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
Running fgsea on the dream results.

```{r}
# Set seed
set.seed(123456)
# Set project library
.libPaths("results/")
library(tidyverse) 
library(RColorBrewer) 
library(fgsea)
# Set relevant paths
list.files()
bg_path <- "data/GSEA_references/Reference gene sets/msigdb_v2023.2.Hs_GMTs/"
out_path <- "results/pt14/results/DREAM"
in_path<-"results/"

# Functions ===================================================
## Function: Adjacency matrix to list -------------------------
matrix_to_list <- function(pws){
  pws.l <- list()
  for (pw in colnames(pws)) {
    pws.l[[pw]] <- rownames(pws)[as.logical(pws[, pw])]
  }
  return(pws.l)
}
## Function: prepare_gmt --------------------------------------
prepare_gmt <- function(gmt_file, genes_in_data, savefile = FALSE){
  
  # Read in gmt file
  gmt <- gmtPathways(gmt_file)
  hidden <- unique(unlist(gmt))
  
  # Convert gmt file to a matrix with the genes as rows and for each go annotation (columns) the values are 0 or 1
  mat <- matrix(NA, dimnames = list(hidden, names(gmt)),
                nrow = length(hidden), ncol = length(gmt))
  for (i in 1:dim(mat)[2]){
    mat[,i] <- as.numeric(hidden %in% gmt[[i]])
  }
  
  #Subset to the genes that are present in our data to avoid bias
  hidden1 <- intersect(genes_in_data, hidden)
  mat <- mat[hidden1, colnames(mat)[which(colSums(mat[hidden1,])>5)]] # filter for gene sets with more than 5 genes annotated
  # And get the list again
  final_list <- matrix_to_list(mat) # for this we use the function we previously defined
  
  if(savefile){
    saveRDS(final_list, file = paste0(gsub('.gmt', '', gmt_file), '_subset_', format(Sys.time(), '%d%m'), '.RData'))
  }
  
  print('Wohoo! .gmt conversion successfull!:)')
  return(final_list)
}
```

# Analysis ====================================================

```{r}
dream<-read.csv("results/Pt14/results/Dream/Pt14_DREAM_LMPP.csv")
#Prepare ranking of your DF gene
rankings<- sign(dream$log2FC)*(-log10(dream$p.value)) #use signed p values as preferred
names(rankings)<-dream$Gene
#Check list
head(rankings)
#Sort genes by signed p value
rankings<-sort(rankings, decreasing= TRUE)
#Plot
plot(rankings)
#Check for infinite values
max(rankings)
min(rankings)
#if there are infinite values adjust them for very high numbers e.g. 10x the highest non-infinite value
#max_ranking <- max(rankings[is.finite(rankings)])
#min_ranking <- min(rankings[is.finite(rankings)])
#rankings <- replace(rankings, rankings > max_ranking, max_ranking * 10)
#rankings <- replace(rankings, rankings < min_ranking, min_ranking * 10)
#rankings <- sort(rankings, decreasing = TRUE) # sort genes by ranking

#Plot first ranked genes of you want (optional)
ggplot(data.frame(gene_symbol = names(rankings)[1:50], ranks = rankings[1:50]), aes(gene_symbol, ranks)) + 
  geom_point() +
  theme_classic() + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```
#With the rankings list we now run fgsea on different backgrounds

#GO

```{r}
my_genes <- dream$Gene
list.files(bg_path)
gmt_files <- list.files(path = bg_path, pattern = '.gmt', full.names = TRUE)
#according to this list KEGG = [14], GO = [53], reactome = [17], wikipathways [21], misigdb [69], this script uses KEGG, GO and Misigdb
#GO pathways
bg_genes_GO <- prepare_gmt(gmt_files[53], my_genes, savefile = FALSE)

#Run GSEA
GSEAres_GO <- fgsea(pathways = bg_genes_GO, # List of gene sets to check
                 stats = rankings,
                 scoreType = 'std', # in this case we have both pos and neg rankings. if only pos or neg, set to 'pos', 'neg'
                 minSize = 10,
                 maxSize = 500,
                 nproc = 1) # for parallelisation
#Check whether it worked
head(GSEAres_GO)
write_csv(GSEAres_GO, "results/pt14/results/DREAM/GSEA_LMPP_pt14_GO_dream.csv")

#Order by pvalue
head(GSEAres_GO[order(pval),])
sum(GSEAres_GO[,padj<0.05])
sum(GSEAres_GO[,pval<0.01])


# plot the most significantly enriched pathway
library(ggplot2)

GSEAres_GO$adjPvalue <- ifelse(GSEAres_GO$padj <= 0.05, "significant", "non-significant") #no pavlues under 0.05
cols <- c("non-significant" = "grey", "significant" = "red")
#GSEAres_GO_sig=subset(GSEAres_GO, GSEAres_GO$padj<=0.1)
ggplot(GSEAres_GO, aes(reorder(pathway, NES), NES, fill = adjPvalue)) +
  geom_col() +
  scale_fill_manual(values = cols) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  coord_flip() +
  labs(x="Pathway", y="Normalized Enrichment Score",
       title="GO pathways Enrichment Score from GSEA")
ggsave("GO_pathways_deseq2_pt14_LMPP.pdf", path = "results/Pt14/results/Figures", width =10, height = 10, unit = "cm")

#Save
#Save the results -----------------------------------------------
data.table::fwrite(GSEAres_GO, file ="results/pt14/results/DREAM/GSEA_pt14_BL_REL_LMPP_GO_dream.tsv" , sep = "\t", sep2 = c("", " ", ""))
```

######KEGG pathways

```{r}
gmt_files <- list.files(path = bg_path, pattern = '.gmt', full.names = TRUE)

#according to this list KEGG = [12], GO = [54], reactome = [18], wikipathways [21], misigdb [69], this script uses KEGG, GO and Misigdb
bg_genes_KEGG <- prepare_gmt(gmt_files[12], my_genes, savefile = FALSE)

#Run GSEA
GSEAres_KEGG <- fgsea(pathways = bg_genes_KEGG, # List of gene sets to check
                 stats = rankings,
                 scoreType = 'std', # in this case we have both pos and neg rankings. if only pos or neg, set to 'pos', 'neg'
                 minSize = 10,
                 maxSize = 500,
                 nproc = 1) # for parallelisation
#Check whether it worked
head(GSEAres_KEGG)
write_csv(GSEAres_KEGG,"results/Pt14/results/DREAM/GSEA_pt14_LMPP_KEGG_dream.csv")

#Order by pvalue
head(GSEAres_KEGG[order(pval),])
sum(GSEAres_KEGG[,padj<0.05])
sum(GSEAres_KEGG[,pval<0.01])


# plot the most significantly enriched pathway
plotEnrichment(bg_genes_KEGG[[head(GSEAres_KEGG[order(GSEAres_KEGG$padj), ], 1)$pathway]],
               rankings) + 
  labs(title = head(GSEAres_KEGG[order(padj), ], 1)$pathway)

#Barplot
GSEAres_KEGG$adjPvalue <- ifelse(GSEAres_KEGG$padj <= 0.05, "significant", "non-significant")
cols <- c("non-significant" = "grey", "significant" = "red")
GSEAres_KEGG_sig=subset(GSEAres_KEGG, GSEAres_KEGG$padj<=0.05)
ggplot(GSEAres_KEGG_sig , aes(reorder(pathway, NES), NES, fill = adjPvalue)) +
  geom_col() +
  scale_fill_manual(values = cols) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  coord_flip() +
  labs(x="Pathway", y="Normalized Enrichment Score",
       title="KEGG pathways Enrichment Score from GSEA (p<0.05)")
ggsave("KEGG_pt14_LMPP_dream.pdf", path = "results/Pt14/results/Figures", width =20, height = 30, unit = "cm")
#Save
data.table::fwrite(GSEAres_KEGG, file ="results/Pt14/results/DREAM/GSEA_pt14_BL_REL_LMPP_KEGG_dream.tsv" , sep = "\t", sep2 = c("", " ", ""))
```
######HALLMARK

```{r}
gmt_files <- list.files(path = bg_path, pattern = '.gmt', full.names = TRUE)
#according to this list KEGG = [14], GO = [53], reactome = [17], wikipathways [21], misigdb [69], this script uses KEGG, GO and Misigdb
bg_genes_HALLMARK <- prepare_gmt(gmt_files[68], my_genes, savefile = FALSE)

#Run GSEA
GSEAres_HALLMARK <- fgsea(pathways = bg_genes_HALLMARK, # List of gene sets to check
                          stats = rankings,
                          scoreType = 'std', # in this case we have both pos and neg rankings. if only pos or neg, set to 'pos', 'neg'
                          minSize = 10,
                          maxSize = 500,
                          nproc = 1) # for parallelisation
#Check whether it worked
head(GSEAres_HALLMARK)
write_csv(GSEAres_HALLMARK,"results/Pt14/results/DREAM/GSEA_pt14_LMPP_HALLMARK_dream.csv")

#Order by pvalue
head(GSEAres_HALLMARK[order(pval),])
sum(GSEAres_HALLMARK[,padj<0.05])
sum(GSEAres_HALLMARK[,pval<0.01])


# plot the most significantly enriched pathway
plotEnrichment(bg_genes_HALLMARK[[head(GSEAres_HALLMARK[order(padj), ], 1)$pathway]],
               rankings) + 
  labs(title = head(GSEAres_HALLMARK[order(padj), ], 1)$pathway)
#Save
data.table::fwrite(GSEAres_HALLMARK, file ="results/Pt14/results/DREAM/GSEA_pt14_BL_REL_LMPP_HALLMARK_dream.tsv" , sep = "\t", sep2 = c("", " ", ""))
```

```{r}
GSEAres_HALLMARK$adjPvalue <- ifelse(GSEAres_HALLMARK$padj <= 0.05, "significant", "non-significant")
cols <- c("non-significant" = "grey", "significant" = "red")
GSEAres_HALLMARK_sig=subset(GSEAres_HALLMARK, GSEAres_HALLMARK$padj<0.05)
ggplot(GSEAres_HALLMARK , aes(reorder(pathway, NES), NES, fill = adjPvalue)) +
  geom_col() +
  scale_fill_manual(values = cols) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  coord_flip() +
  labs(x="Pathway", y="Normalized Enrichment Score",
       title="Hallmark pathways Enrichment Score from GSEA")
ggsave("HALLMAR_pt14_LMPP_Deseq_dream.pdf", path = "results/Pt14/results/Figures", width =25, height = 30, unit = "cm")
```
```{r}
head(GSEAres_HALLMARK)
```

#All pathways

```{r}
gmt_files <- list.files(path = bg_path, pattern = '.gmt', full.names = TRUE)
#according to this list KEGG = [14], GO = [53], reactome = [17], wikipathways [21], misigdb [69], this script uses KEGG, GO and Misigdb
bg_genes_all <- prepare_gmt(gmt_files[5], my_genes, savefile = FALSE)

#Run GSEA
GSEAres_all <- fgsea(pathways = bg_genes_all, # List of gene sets to check
                          stats = rankings,
                          scoreType = 'std', # in this case we have both pos and neg rankings. if only pos or neg, set to 'pos', 'neg'
                          minSize = 10,
                          maxSize = 500,
                          nproc = 1) # for parallelisation
#Check whether it worked
head(GSEAres_all)
write_csv(GSEAres_all,"results/Pt14/results/DREAM/GSEA_pt14_LMPP_pathways_dream.csv")

#Order by pvalue
head(GSEAres_all[order(pval),])
sum(GSEAres_all[,padj<0.05])
sum(GSEAres_all[,pval<0.01])


# plot the most significantly enriched pathway
plotEnrichment(bg_genes_all[[head(GSEAres_all[order(padj), ], 1)$pathway]],
               rankings) + 
  labs(title = head(GSEAres_all[order(padj), ], 1)$pathway)
#Save
data.table::fwrite(GSEAres_all, file ="results/Pt14/results/DREAM/GSEA_pt14_BL_REL_LMPP_all_dream.tsv" , sep = "\t", sep2 = c("", " ", ""))

```

```{r}
#Visualise
GSEAres_all$adjPvalue <- ifelse(GSEAres_all$padj <= 0.05, "significant", "non-significant")

cols <- c("non-significant" = "grey", "significant" = "red")
GSEAres_all_sig=subset(GSEAres_all, GSEAres_all$padj<0.05)
ggplot(GSEAres_all_sig , aes(reorder(pathway, NES), NES, fill = adjPvalue)) +
  geom_col() +
  scale_fill_manual(values = cols) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  coord_flip() +
  labs(x="Pathway", y="Normalized Enrichment Score",
       title="Allpathways Enrichment Score from GSEA (p<0.05)")
ggsave("allPathways_pt14_LMPP_Dream.pdf", path = "results/Pt14/results/Figures", width =20, height = 35, unit = "cm")
```

For other visualisation this code will make a list which is compatible with clusterprofiler.

```{r cars}
# For data management
library('tidyverse')
library("clusterProfiler")
library("org.Hs.eg.db")
# For visualisation
library('pheatmap')
library("DOSE")
library("enrichplot")
library("ggupset")
```

Define function

```{r}
# Functions ------------------------------------------------
## Function: Adjacency matrix to list ####
matrix_to_list <- function(pws){
  pws.l <- list()
  for (pw in colnames(pws)) {
    pws.l[[pw]] <- rownames(pws)[as.logical(pws[, pw])]
  }
  return(pws.l)
}
```

Annotate

```{r}
dream<-dream%>%mutate(diffexpr = case_when(
  log2FC>0 &adjusted.pval<0.05 ~'UP',
  log2FC<0 &adjusted.pval<0.05 ~'DOWN',
  adjusted.pval>0.05 ~"NO"
))
```

```{r}
head(dream$diffexpr)
```

```{r}
dream<-dream[dream$diffexpr!="NO",]
```

Prepare the background gene set

```{r}
list.files(bg_path)
bg_genes_HALLMARK <- read.gmt("data/GSEA_references/Reference gene sets/msigdb_v2023.2.Hs_GMTs/h.all.v2023.2.Hs.symbols.gmt")
genes_in_data<-dream$Gene
bg_genes_all<-read.gmt("data/GSEA_references/Reference gene sets/msigdb_v2023.2.Hs_GMTs/c2.all.v2023.2.Hs.symbols.gmt")
head(bg_genes_all)
```

```{r}
deg_results_list<-split(dream, dream$diffexpr)
```

```{r}
padj_cutoff <-0.05
genecount_cutoff<-5
genes_in_data<-dream$Gene
nrow(df)
nrow(bg_genes_HALLMARK)
bg_genes <- read.gmt("data/GSEA_references/Reference gene sets/msigdb_v2023.2.Hs_GMTs/h.all.v2023.2.Hs.symbols.gmt")
bg_genes<-bg_genes[bg_genes$gene%in%deseq$gene,]
```

```{r}
# Run clusterProfiler on each sub-dataframe
res <- lapply(names(deg_results_list),
                 function(x) enricher(gene = deg_results_list[[x]]$Gene,
                                      TERM2GENE =  bg_genes
                                      ))
names(res) <- names(deg_results_list)

#Convert the list of enrichResults for each sample_pattern to a dataframe with the pathways
res_df <- lapply(names(res), function(x) rbind(res[[x]]@result))
names(res_df) <- names(res)
res_df <- do.call(rbind, res_df)
head(res_df)

```

Make a nicer table

```{r}
#Convert the enrichResults to a dataframe with the pathways
res_df <- lapply(names(res), function(x) rbind(res[[x]]@result))
names(res_df) <- names(res)
res_df <- do.call(rbind, res_df)
head(res_df)
```

Filter the output

```{r}
target_pw<-unique(res_df$ID[res_df$p.adjust<padj_cutoff&res_df$Count>genecount_cutoff])
res_df<-res_df[res_df$ID%in%target_pw,]
```

```{r}
enrichres_Hallmark <- new("enrichResult",
                 readable = FALSE,
                 result = res_df,
                 pvalueCutoff = 0.05,
                 pAdjustMethod = "BH",
                 qvalueCutoff = 0.2,
                 organism = "human",
                 ontology = "UNKNOWN",
                 gene = deseq$gene,
                 keytype = "UNKNOWN",
                 universe = unique(bg_genes_HALLMARK$gene),
                 gene2Symbol = character(0),
                 geneSets = bg_genes)
class(enrichres_Hallmark)
```

Barplot

```{r}
barplot(enrichres_Hallmark, showCategory = 20) 
mutate(enrichres_Hallmark, qscore = -log(p.adjust, base = 10)) %>% 
  barplot(x = "qscore")
```

Dotplot

```{r}
enrichplot::dotplot(enrichres_Hallmark, x = "geneRatio",showCategory = 15, orderBy = "p.adjust") 
ggsave("Hallmark_pt14_LMPP_Dream_dotplot.pdf", path = "results/Pt14/results/Figures", width =20, height = 35, unit = "cm")
```

and for all pathways

```{r}
deg_results_list<-split(dream, dream$diffexpr)
```

```{r}
padj_cutoff <-0.05
genecount_cutoff<-5
genes_in_data<-dream$Gene
nrow(df)
nrow(bg_genes_HALLMARK)
bg_genes <- read.gmt("data/GSEA_references/Reference gene sets/msigdb_v2023.2.Hs_GMTs/c2.all.v2023.2.Hs.symbols.gmt")
bg_genes<-bg_genes[bg_genes$gene%in%deseq$gene,]
```

```{r}
# Run clusterProfiler on each sub-dataframe
res <- lapply(names(deg_results_list),
                 function(x) enricher(gene = deg_results_list[[x]]$Gene,
                                      TERM2GENE =  bg_genes
                                      ))
names(res) <- names(deg_results_list)

#Convert the list of enrichResults for each sample_pattern to a dataframe with the pathways
res_df <- lapply(names(res), function(x) rbind(res[[x]]@result))
names(res_df) <- names(res)
res_df <- do.call(rbind, res_df)
head(res_df)
```

```{r}
#Convert the enrichResults to a dataframe with the pathways
res_df <- lapply(names(res), function(x) rbind(res[[x]]@result))
names(res_df) <- names(res)
res_df <- do.call(rbind, res_df)
head(res_df)
```

```{r}
target_pw<-unique(res_df$ID[res_df$p.adjust<padj_cutoff&res_df$Count>genecount_cutoff])
res_df<-res_df[res_df$ID%in%target_pw,]
```

```{r}
enrichres_all <- new("enrichResult",
                 readable = FALSE,
                 result = res_df,
                 pvalueCutoff = 0.05,
                 pAdjustMethod = "BH",
                 qvalueCutoff = 0.2,
                 organism = "human",
                 ontology = "UNKNOWN",
                 gene = deseq$gene,
                 keytype = "UNKNOWN",
                 universe = unique(bg_genes_all$gene),
                 gene2Symbol = character(0),
                 geneSets = bg_genes)
class(enrichres_all)
```

```{r}
barplot(enrichres_all, showCategory = 20) 
mutate(enrichres_all, qscore = -log(p.adjust, base = 10)) %>% 
  barplot(x = "qscore")
```

```{r}
enrichplot::dotplot(enrichres_all, x = "geneRatio",showCategory = 25, orderBy = "GeneRatio") 
ggsave("Hallmark_pt14_LMPP_Dream_dotplot_all.pdf", path = "results/Pt14/results/Figures", width =20, height = 35, unit = "cm")
```