---
title: "Pt9_LMPP_MPP_MPP_DGE"
output: html_document
date: "2024-06-16"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
library(Seurat)
library(tidyverse)
```

By projecting the sorted cells of patient 9, we have identified LMPP_MPP-like population where the major clone mostly projects to.
Load the data and select Pt9, dominant clone for BL and REL and LMPP_MPP-like population.
```{r}
df<-readRDS("data/4_Seurat_obj_CNA_referenceannotations.rds")
#Create a subset for patient 9 BL and REL
pt9<-subset(x=df, subset=Patient=='pt9')
#Extract metadata
data9<-pt9@meta.data
#Create variable marking the dominant clone at BL and REL
table(data9$Sample, data9$clone.y)
data9<-mutate(data9, dominant = if_else(
  pt_status == 'pt9 BL'&clone.y %in% c("AI_5_11")|
    pt_status == 'pt9 REL' & clone.y == 'AISF', "1","0"))
#Check whether it worked
table(data9$dominant, data9$clone.y, data9$Sample)
#Add metadata back to Seurat object
pt9<-AddMetaData(object = pt9, metadata = data9, col.name = 'dominant')
pt9_BL_REL<-subset(x=pt9, pt_status %in% c("pt9 BL","pt9 REL"))
#This is the first analysis, so we have to determine the mainly expanded populations first
table_pt9<-table(pt9_BL_REL$predicted_CellType, pt9_BL_REL$clone.y, pt9_BL_REL$pt_status)
write.csv(table_pt9, "results/pt9_clones_per_celltype.csv")
#According to this table the major clone at BL and at REL is mainly found in LMPP_MPP and MPP, as well as a bit in HSC
#Try with combining MPP and LMPP_MPP
pt9_BL_REL<-subset(x=pt9_BL_REL, subset=predicted_CellType%in% c("LMPP_MPP", "MPP-MkEry", "MPP-MyLy"))
##From this subset select the dominant clones
pt9_BL_REL <- subset(x=pt9_BL_REL, dominant == "1")
#Check whether the subsetting worked
table(pt9_BL_REL$predicted_CellType)
table(pt9_BL_REL$dominant)
table(pt9_BL_REL$predicted_CellType, pt9_BL_REL$Sample)

```
You should get 699 cells from that and combine MPP and LMPP_MPP.
Deseq pseudobulk workflow
Load necessary libraries
```{r}
library(ExperimentHub)
library(DESeq2)
library(tidyverse)
library(RColorBrewer)
```

Pt9 LMPP_MPPs are compared BL against relapse, logFC schrinkage using apeglm and filtered and unfiltered table are stored.
```{r}
# 1. counts matrix - sample level
# counts aggregate to sample level
pt9_BL_REL$samples <- paste0(pt9_BL_REL$Sample, pt9_BL_REL$Sample_well)

#Create dataset with counts and genes
cts <- AggregateExpression(pt9_BL_REL, 
                           group.by = c("samples"),
                           assays = 'RNA',
                           slot = "counts",
                           return.seurat = FALSE)

cts <- cts$RNA

# transpose
cts.t <- t(cts)

# convert to data.frame
cts.t <- as.data.frame(cts.t)

# get values where to split cell type and sample
splitRows <- gsub('_.*', '', rownames(cts.t))

# fix colnames and transpose
cts.t.modified <- t(cts.t)
cts.t.modified[(1:10), (1:10)]

```


```{r}
# Create Count matrix for DE
#Analyse for all cells
# 1. generate sample level metadata
colData_all <- data.frame(samples = colnames(cts.t.modified))

#2. generate a variable for the condition to compare (e.g. relapse and BL)
colData_all <- colData_all %>%
  mutate(condition = ifelse(grepl('BL', samples), 'baseline', 'relapse')) 

```

Next, we run the actual Deseq2
```{r}
# 1.Create DESeq2 object   
dds_all <- DESeqDataSetFromMatrix(countData = cts.t.modified,
                                  colData = colData_all,
                                  design = ~ condition)

# 2.filter -> optional to exclude genes with a low number of reads
keep <- rowSums(counts(dds_all)) >=10
dds_all <- dds_all[keep,]

# 3. run DESeq2
dds_all <- DESeq(dds_all)

# 4. Check the coefficients for the comparison
resultsNames(dds_all)


# 5. Generate results object
res_all <- results(dds_all, contrast = c("condition", 'relapse', 'baseline'))
res_all
res_all<-as.data.frame(res_all)
```
As recommended we shrink the LFC with apeglm (which is the recommended method)
```{r} 
#6. Shrink the logFC
library(apeglm)
res_ape<-lfcShrink(dds_all, coef = 2, type = "apeglm")


#7.Create a table
res_tbl_all <-res_ape%>% 
  data.frame()%>%
  rownames_to_column(var="gene") %>% 
  as_tibble()
res_tbl_all
#Save the table
write.csv(res_tbl_all,"results/pt9/results/Deseq2/DEA_pt9_LMPP_MPP.csv" )


#8.Filter for only significant Genes
#Set threshold
padj_cutoff <- 0.05
#Subset for signifcant results 
sig_res_all<- subset(res_tbl_all, res_tbl_all$padj<0.05)
write.csv(sig_res_all, "results/Pt9/results/Deseq2/DEA_pt9_LMPP_MPP_padj.csv")
```

As a complementary approach, we are using DREAM to identify DEGs.
```{r}
#Get the metadata table
metadata<-pt9_BL_REL@meta.data
metadata$Cell<-rownames(metadata)
REL<-metadata[which(metadata$Sample == "REL"),]
BL<-metadata[which(metadata$Sample == "BL"),]
#Extract the count matrix
cts<-pt9_BL_REL@assays$RNA$counts
cts <- as.matrix(cts)
#Filter out ERCC
ERCC.genes<-grep("^ERCC-", rownames(cts), value = T)
#-> are not included anymore
# This code is redundant
# This code is redundant
cellsREL.m<-cts[,colnames(cts) %in% row.names(REL)]
cellsBL.m<-cts[,colnames(cts) %in% row.names(BL)]

MeanREL<-Matrix::rowMeans(cellsREL.m)
MeanBL<-Matrix::rowMeans(cellsBL.m)
Foldchange<-MeanREL/MeanBL

cellsREL.bi <- cellsREL.m
cellsREL.bi[cellsREL.bi >0] <- 1
REL.freq<-Matrix::rowSums(cellsREL.bi)

cellsBL.bi <- cellsBL.m


cellsBL.bi[cellsBL.bi >0] <- 1
BL.freq<-Matrix::rowSums(cellsBL.bi)

ExpFractionREL=REL.freq/ncol(cellsREL.m)
ExpFractionBL=BL.freq/ncol(cellsBL.m)

z0<-data.frame(Gene=rownames(cts),
               ExpREL=MeanREL,
               ExpBL=MeanBL,
               FoldChange=Foldchange,
               ExpFreqREL=REL.freq,
               ExpFreqBL=BL.freq,
               TotalREL=ncol(cellsREL.m),
               TotalBL=ncol(cellsBL.m),
               ExpFractionREL=ExpFractionREL,
               ExpFractionBL=ExpFractionBL)
rm(cellsREL.m, cellsBL.m, cellsREL.bi, cellsBL.bi)
````

```{r}
# Filter out genes expressed in fewer cells than min.expFraction
z0<-subset(z0,ExpFractionREL >=0.1 | ExpFractionBL >=0.1)
cts<-cts[rownames(cts) %in% as.character(z0$Gene),]

# Normalize using scran size factors
#this is  absolutely necessary
# Get scran size factors
sf <- metadata$sizeFactor

nsf <- log(sf/Matrix::colSums(cts))
nsf <- exp(nsf - mean(nsf, na.rm=T))

dge <- edgeR::DGEList(counts = cts,
                      lib.size = Matrix::colSums(cts),
                      norm.factors = nsf,
                      remove.zeros = FALSE)

#Make a conditions dataframe
conditions.df <- data.frame(Cell = metadata$Cell,
                            condition = metadata$Sample,
                            covariate = metadata$Sort) %>%
  column_to_rownames("Cell")

```
Only now we can run dream
```{r}
param = BiocParallel::SnowParam(2, "SOCK", progressbar=TRUE)

# estimate weights using linear mixed model of dream
print("Estimating weights using linear mixed model of dream")
formula <- ~ Sample
vobjDream = variancePartition::voomWithDreamWeights( dge, ~condition, conditions.df , BPPARAM=param )


# Fit the dream model on each gene
# By default, uses the Satterthwaite approximation for the hypothesis test
print("Fitting dream model on each gene")
fitmm = variancePartition::dream( vobjDream, ~condition, conditions.df , BPPARAM=param )
fitmm = variancePartition::eBayes(fitmm)
```

```{r}
# Examine design matrix
print(head(fitmm$design, 3))

# Get results of hypothesis test on coefficients of interest
resT <- variancePartition::topTable( fitmm, coef=2, number=Inf )
resT <- tibble::rownames_to_column(resT, "Gene")

resT <- resT %>% dplyr::rename(p.value = P.Value, 
                               adjusted.pval = adj.P.Val, 
                               log2FC = logFC)
z0 <- dplyr::left_join(z0, resT,by = "Gene")

z0$ranking<- (-log10(z0$p.value))*sign(z0$log2FC)
z0<-z0 %>% dplyr::arrange(-ranking)

z0 %>%
  ggplot(aes(x = FoldChange, y = log2FC)) +
  geom_point()

write_csv(z0, "results/Pt9/results/DREAM/Pt9_DREAM_LMPP_MPP.csv")
```


